---
title: "7784 Skills: Programming in R"
author: 'Kirill Müller, Christoph Sax'
date: 'University of St. Gallen, 14.11.2019'        # 1\. to avoid itemizing
output:
  cynkradown::cynkra_slides:
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    selfcontained: true
fontsize: 10pt
lang: english                # ngerman, english, italian, french
font: frutiger
wide: false
colorlinks: false
logo: false
header-includes:
  - \usepackage{parskip}
---

```{r setup, include = FALSE}
library(tidyverse)
set.seed(20191010)

ggplot <- function(...) ggplot2::ggplot(...) + theme_bw(16)
```

class: middle, inverse

# 14.11.: Visualizing statistical output


---
# Topics for today

- Homework 5, 6

- Two presentations

- Visualizing statistical output

- More on linear regression


---
# Schedule

  Date       | First Slot | Second Slot
-------------|------------|------------
  10/24/2019 | team04     | team10
  11/14/2019 | team09     | team03
  11/21/2019 | team12     | team01
  11/28/2019 | team07     | team05
  12/05/2019 | team06     | team11
  12/12/2019 | team08     |
  12/19/2019 |            |


---


class: middle, inverse

# Homework (so far)


---

# Do not leave interactive functions in your code

- no `view()`, `View()`

- no `install.packages()`


---


# Ensure reproducibility

- Restart R (Shift + Ctrl + F10)

- Run script (Shift + Ctrl + Enter)

- Make sure there are no errors


---


# Style

- https://style.tidyverse.org/

- Use the [styler](https://cran.r-project.org/web/packages/styler/index.html) package


---

# Style II
Most infix operators (`==`, `+`, `-`, `<-`, etc.) should always be surrounded by spaces:


```r
# Good
height <- (feet * 12) + inches
mean(x, na.rm = 10)

# Bad
height<-feet*12+inches
mean(x, na.rm=10)
```

---
# Style III


```r
# Good
do_something_very_complicated(
  something = "that",
  requires = many,
  arguments = "some of which may be long"
)

# Bad
do_something_very_complicated("that", requires, many, arguments,
                              "some of which may be long"
                              )
```

---
# Style IV

Use `TRUE` not `T`, because:

```{r, error=TRUE}
TRUE <- 1
FALSE <- 0

# But
T <- F
```
---


# Style V

The `c()` function is only needed to concatenate vectors. Do not write it otherwise.

Needed:
```r
c(5, 5)
```

Not needed:
```r
c(rnorm(5))

rnorm(5)  # better!
```

---

# `lesson05.R`, first exercise

In the matrix section, questions 6 and 7 wrongly referred to questions 3 and
4, rather than to questions 4 and 5, and made not much sense. They should
have been stated as:

- Compare the classes of the results in 4. and 5. to each other.

- Modify 4., so that it returns the same class as 5.

---

# `lesson06.R`, last exercise

> Add a dummy variable, `DCATH`, that contains 0 if the share of Catholics
is below 50, 1 otherwise. Verify that you get the same result as if you
directly included a categorical variable.

Dummy variable:
```r
c(1, 1, 1, 1, 0, 0)
```

Categorical variable
```r
c("catholic", "catholic", "catholic", "catholic", "protestant", "protestant")
```










