# Basics of R base

[Download as R script](05-r-base.R)


[Intro Slides](05-r-base.html)

We have spent the first lessons of this course with dplyr and ggplot2, two of
the great innovations in R universe of the last few years. If you were
learning R 10 years ago, R would have looked quite differently. In this
lesson, we cover the basics of R base - *R (mostly) as it is before you load
any packages*.

## The Main Data Structures

R has four main data structures. In dplyr, we always work on **data frames**,
or **tibbles**, the equivalent to an Excel sheet - a tabular collection of
data. When performing operations, e.g., by using `filter()`, or `mutate()` we
are performing operations on the columns of a data frame. These columns are
**vectors**, and they are the heart of R. Related to them, **matrices** are
basically two dimensional vectors that are useful in some mathematical or
statistical applications. Finally, **lists** are the most versatile data
type, and they are used in many circumstances. We will cover these data
structures in the following order: vectors, matrices, lists and data frames,
or tibbles (which we already know well).


### Vectors, the R Workhorse

The vector type is the most basic data structure in R. It's hard to imagine R
code that doesn't involve vectors. If you use `mutate` to change columns of
your data frame, you are effectively operating on vectors. The elements of a
vector must all have the same class, or data type. You can have a vector
consisting of three character strings (of class character) or three integer
elements (of class integer), but not a vector with one integer element and
two character string elements.


#### Vector Classes

In many programming languages, vector variables are considered different from
scalars, which are single-number variables. However, in R, numbers are
actually considered one-element vectors, and there is really no such thing as
a scalar.

As we have seen previously, all elements of an R vector (or data frame
column) must have the same class, which can be integer, numeric (also called
double), character (string), logical (boolean).  Here are some examples:

```{r }
x_log <- c(TRUE, FALSE)     # same as c(T, F)
x_int <- c(1L, 2L, 3L)      # use 1L to enforce integer, rather than numeric
x_num <- c(1, 2, 6.3)       # also called 'double'
x_chr <- c("Hello World")
```

If you need to check the class of a variable `x`, you can use, e.g.:

```{r }
class(x_log)
```

There is a certain order in the list above: `logical` is the least flexible
class, while `character` is the most flexible. If you combine vectors of
different classes, the more flexible class will win:

```{r }
class(c(x_log, x_num))
class(c(x_int, x_chr))
```

You can change the class of a vector with the following coercion functions.

```{r }
as.logical(c(1, 0))
as.integer(c(1, 0))
as.numeric(c("1", "2"))
as.character(c(TRUE, FALSE))
```

These functions will always work if you coerce towards greater flexibility.
If you want to go the other way, it may give you `NA`s and some warnings:

```{r }
as.numeric(c("hi", "number", "1"))
```

#### Recycling

When applying an operation to two vectors that requires them to be the same
length, R automatically recycles, or repeats, the shorter one, until it is
long enough to match the longer one. Here is an example:

```{r }
c(1, 2) + c(6, 0, 9, 20, 22, 11)
```

The shorter vector was recycled, so the operation was taken to be as follows:

```{r }
c(1, 2, 1, 2, 1, 2) + c(6, 0, 9, 20, 22, 11)
```

The most common recycling operation involves a vector of length 1:

```{r }
c(6, 0, 9, 20, 22, 11) + 3
```

In dplyr, only a vector of length 1 is allowed to recycle, the other cases
will result in an error.

```{r }
# tibble(a = c(1, 2), b = c(6, 0, 9, 20, 22, 11))
```

#### Arithmetic Operators

In R, every operator, including `+` in the following example, is actually a
function.

```{r }
2 + 3
```

The `+` here is a function with two arguments. A more functional way of
writing it is the following:

```{r }
"+"(2, 3)
```

Remember that scalars are actually one-element vectors. So, we can add
vectors, and the `+` operation will be applied element-wise.

```{r }
x <- c(1, 2, 4)
x + c(5, 0, -1)
```

The same is true, e.g., for multiplication, which is done element by element
as well. (We will have a look at matrix multiplication in the next section.)

```{r }
x * c(5, 0, -1)
```

#### Comparison Operators
Similar to arithmetic operators, comparison operators are applied element
wise. The following expression will return a single `TRUE`, as we are
comparing two vectors of length 1:

```{r }
2 > 1
```

The comparison operator for 'is equal to' is `==`, not `=`:

```{r }
1 + 1 == 2
```

Here is how they work on longer vectors:

```{r }
x <- c(1, 2, 4, 2)
y <- c(2, 2, 4, 5)
x == y
```

The usual recycling rules apply as well:

```{r }
x == 2
```

Here are the other comparison operators:

  - `x < y`: less than
  - `x <= y`: less or equal than
  - `x >= y`: greater or equal than
  - `x != y`: not equal
Logical vectors can be combined by `&` (AND) or `|` (OR):

```{r }
a <- x >= 2
b <- x < 4
a & b

x >= 4 | x < 2
```

#### Indexing

One of the most frequently used operations in R base is that of indexing
vectors, in which we form a subvector by picking elements. You can use both
integer values or logical  vectors for indexing.

##### Indexing Using Integers

We can extract values from a vector, using an integer index:

```{r }
y <- c(1.2, 3.9, 0.4, 0.12)
y[c(1, 3)]
y[2:3]
v <- 3:4
y[v]
```

Note that duplicates are allowed:

```{r }
y[c(1, 1, 3)]
```

Negative subscripts mean that we want to exclude the elements:

```{r }
z <- c(5, 12, 13)
z[-1]
z[-1:-2]
```

##### Logical Indexing

Logical indexing is perhaps even more important. Building on the example from
above, we could also select element 1 and 3 in the following way:

```{r }
y <- c(1.2, 3.9, 0.4, 0.12)
y[c(FALSE, TRUE, FALSE, TRUE)]
```

Logical indexing picks the `TRUE`s but not the `FALSE`s. This is the main
building block for filtering. Suppose you have:

```{r }
y[y > 1]
```

This will return all elements that are bigger than 1. How is this done?
First, R had evaluated the comparison, `y > 1`, which led to logical vector:

```{r }
y > 1
```

Second, using logical indexing, this vector was then used to pick those
elements that evaluated to `TRUE`. So `y[y > 1]` is actually the same as:

```{r }
y[c(TRUE, TRUE, FALSE, TRUE)]
```

We have already seen that you can assign to individual elements of a vector,
using integer indices:

```{r }
y[c(2, 4)] <- 5
```

Of course, you can do the same with logical indices:

```{r }
y[c(FALSE, TRUE, FALSE, TRUE)] <- 5
```

This is a very powerful tool. For example, if you want to truncate all
negative numbers in a vector to 0, you can use:

```{r }
z <- c(-3, 1.2, 2, -22)
z[z < 0] <- 0
```

Or use it with the `%in%` operator you encountered before:

```{r }
z[z %in% y]
```

#### Exercises

1. Create a vector called `v1` containing the numbers 2, 5, 8, 12 and 16.

2. Extract the values at positions 2 and 5 from `v1`.

3. Use `x:y` notation to make a second vector called `v2` containing the
numbers 5 to 9.

4. Subtract `v2` from `v1` and look at the result.

5. Generate a vector with 1000 standard-normally distributed random numbers
(use `rnorm()`). Store the result as `v3`. Extract the numbers that are
bigger than 2.


### Matrices

An R matrix corresponds to the mathematical concept of the same name: a
rectangular array of numbers (most of the time), or some other type.
Here is some sample matrix code:

```{r }
m <- matrix(c(1, 4, 2, 2), nrow = 2, ncol = 2)
m
```

The main use of matrices is for matrix algebra. You can do all kind of matrix
algebra operations, right out of R base:

```{r }
m %*% m         # matrix multiplication
m * m           # elementwise multiplication

m %*% solve(m)  # inverse of a matrix
m / m           # elementwise division

m + m
m - m

t(m)            # matrix transpose
qr(m)           # QR decomposition
det(m)          # determinant
eigen(m)        # eigenvalues/eigenvectors
diag(m)         # diagonal
```

Matrices are indexed by double subscripting:

```{r }
m[1, 2]
m[2, 2]
```

You can extract submatrices from a matrix, much as you extract subvectors
from vectors:

```{r }
m[1, ]   # row 1
m[1, , drop = FALSE]  # keeps being a matrix

m[, 2]  # column 2
```

#### Exercises

1. Create a 10 x 10 matrix that contains a sequence of numbers (use the `:`
notation).

2. Use the transpose function on the matrix

3. Extract the 2. column of the matrix

4. Extract the 5. row of the matrix

5. Extract the 5. and the 6. row of the matrix

6. Compare the classes of the results in 3. and 4. to each other

7. Modify 3., so that it returns the same class as 4.


### Lists

Like an R vector, an R list is a container for values, but its contents can
be items of different data types, or different length. Here's an example:

```{r }
x <- list(u = c(2, 3, 4), v = "abc")
x

x$u
```

The expression `x$u` refers to the `u` component in the list `x`.

```{r }
x[['u']]
x[[1]]
```

We can also refer to list components by their numerical indices. However,
note that in this case, we use double brackets instead of single ones.

We can also use single brackets rather than double brackets to get a subset
of the list.

```{r }
x['u']
x[1]
x[1:2]
```

Note that `x[[1]]` returns the component (a numeric vector), while `x[1]`
returns a subset of the list (a list of length 1):

```{r }
class(x[1])
class(x[[1]])
```

Hadley Wickham's visualization helps a lot here:

https://twitter.com/hadleywickham/status/643381054758363136
Lists are not restricted to containing vectors. In fact, they can contain
anything, for example, a data frames:

```{r }
ll <- list(mtcars = mtcars, u = c(2, 3, 4))
```

#### Exercises

1. Generate two random vectors of length 10, `a`, and `b`. Combine them in a
list, call it `l1`.

2. Compare the classes of `l1[2]` and `l1[[2]]`. Can you explain the
difference?


### Data Frames

As we saw in many places, a typical data set contains data of different
classes.  Instead of a matrix, we use a data frame, or tibble. Technically, a
data frame in R is a list, with each component of the list being a vector
corresponding to a column in our data. You can create a data frame using
`tibble` from tidyverse:

```{r }
library(tidyverse)
d <- tibble(kids = c("Jack", "Jill"), ages = c(12, 10))
d
```

Here, I am using the `tibble()` function from tidyverse, rather than the R
base equivalent, `data.frame()`. The reason for this is that `data.frame()`
has some undesirable features, such as converting character vectors into
factors.  Because data frames are technically a list, we can access its
vectors the same way as we access components of a list.

```{r }
d$ages
d[['ages']]
d[[2]]  # usually not recommended
```

Typically, though, data frames are created by reading in a data set from a
file or a database, as we saw in the previous days of the workshop.

Contrary to lists, data frames require their columns to be of the same
length. If they are not, values will be recycled. That is why the following
works:

```{r }
tibble(kids = c("Jack", "Jill"), ages = c(12, 10), type = "kid")
```

## Introduction to Functions

As in most programming languages, the heart of R programming consists of
writing functions. A function is a group of instructions that takes inputs,
uses them to compute other values, and returns a result.

Let's write a function that divides all elements of a vector by 2:

```{r }
half <- function(x) {
  x / 2
}
```

This is a function named `half`, whose purpose is to divides every element of
a vector by 2. It's a pretty pointless function, as the operation itself is
so simple. Arguments to a function are enclosed by parentheses (`()`); the
body of the function is enclosed by braces (`{}`).

Let's see how it works:

```{r }
half(c(3, 2, 1))
half(AirPassengers)
```

here we are saving the output in a variable inside the function:

```{r }
half <- function(x) {
  z <- x / 2
  z
}
```

An R functions will return the last value computed if there is no explicit
`return()` call. We could have been more explicit, but it is usually not
necessary:

```{r }
half <- function(x) {
  z <- x / 2
  return(z)
}
```

Let's make it a bit more complex, by adding an additional argument:

```{r }
fraction <- function(x, denominator){
  x / denominator
}

fraction(c(2, 3, 4), 4)
```

if you have more than one argument, it is a good practice to name the
arguments

```{r }
fraction(x = c(2, 3, 4), denominator = 4)
```

that way, they become independent of the order, which is a very useful if the
number of argument becomes large.

```{r }
fraction(denominator = 4, x = c(2, 3, 4))
```

### Variable Scope
A variable that is visible only within a function body is said to be local to
that function. In `square()`, `x` is a local  variable. In `fraction()`, `x`
and `p` are local variables. They disappear after the function returns:

```{r }
fraction(x = c(2, 3, 4), denominator = 4)

# denominator
# Error: object 'denominator' not found

x
```

Note that the `x` is not the `x` we used in the `fraction` function, but
rather the `x` defined earlier. `x` here is called a global variable, while
the `x` in the function is a local variable. Global variables are shown in
RStudio in the environment pane.  A global variable can be written to from
within a function by using R's superassignment operator, `<<-`, but this is
rarely recommended.


### Default Arguments

A nice feature of R functions is that you can set defaults to arguments.
Let's modify the `fraction` function from above:

```{r }
fraction <- function(x, denominator = 2) {
  x / denominator
}
```

Here `denominator` will be initialized to 2 if the user does not
specify `p` in the
call. So we can use `fraction()` the same way as `half()`:

```{r }
fraction(c(2, 2))
half(c(2, 2))
```

or use its extended capabilities:

```{r }
fraction(c(2, 2), 3)
```

### Exercises

1. Write a function `add_constant` that adds a constant to a vector, and set
the default value of the constant to 10.

2. Apply it to the `AirPassengers` series.

3. You can use your function within dplyr. Using the `mpg` dataset, use
`add_constant` within `mutate` to add a constant value (`100`) to the number
of `cyl`.

4. (evil) We saw that `"+"` is actually a function. In R, it is easily
possible to change the working of such a fundamental function (this is, in
fact, what ggplot does). In order to do so, let's write a new function with
the same name, `"+"`, and two arguments, `a`, and `b`. But instead of summing
the values, let's subtract them (or figure out something more evil).
Verify the result of 1 + 1. Cool, isn't it? (`rm("+")` will restore sanity.)


## Higher-Order Functions

Because functions in R are objects like any other objects, it is easy to
write functions that return functions or take functions as arguments. We saw
some examples above. The most prominent of these functions are *higher-order
functions*, which are very central to R. They are called `apply` or `lapply`,
or similar, and there are many of them. To avoid too much confusion, we will
restrict ourself to `map`, the tidyverse equivalent of `lapply`,


### `map`

The `map` function 'maps' a function to each component of a list.
Because lists are such a useful container for objects in R, very often, you
want to map a function to each component of a list. Here is an example:

```{r }
ll <- list(a = c(2, 3, 4), b = c(1, 2, 3), c = c(5, 2, 1))
```

This is a list with three vectors, of which we want to calculate the means.
Here's an expression that does what we want (we will cover loops later on):

```{r }
z <- NULL
for (vi in ll){
  z <- c(z, mean(vi))
}
```

We loop through each component of the list, calculate the mean and add it to
an output vector, `z`. With the `map` function, this can be written much
more concise:

```{r }
library(tidyverse)
map(ll, mean)
```

So `map` returns a list of the same length as the first argument (a list),
each element of which is the result of applying the second argument (a
function) to the corresponding component of the list argument. If the input
list is named (as in the example), so will be the output.

Additional arguments to the `mean()` function can be included as well:

```{r }
map(ll, mean, na.rm = TRUE)
```

If you want to convert the list to a vector (as in the loop example), use
`unlist` on the result:

```{r }
unlist(map(ll, mean))
```

Of course, you can use `map` with your own functions. Here we want to pick
the second element of each vector in the list:

```{r }
pick_second <- function(x){
  x[2]
}

map(ll, pick_second)
```

In cases like this, we may want to use the possibility of having *anonymous
functions*, i.e. functions without a name. So we just substitute `pick_second`
by its definition:

```{r }
map(ll, function(x) x[2])
```

For simple one line functions like this, it may be justified to omit the
curly braces.
### Exercises

1. Use `map` to calculate the mean of each variable in the `swiss`
dataset. Convert the resulting list to a vector.

2. Use `map` to coerce the variables in the `swiss` dataset to character.

3. Using `map`, generate a list containing 10 random vectors of random
length between 1 and 10.

4. Use the help to see what the `colSums()` function does. Using `apply`, try
writing your own version, `colSums2()`.


### Loops

In many programming languages, one of first things you learn are loops. There
is a reason that we didn't cover them until now, and also only for
completeness. Loops in R are slow, and they are -- most of the time --
unnecessary. This is because many operations in R are vectorized anyway, so
there is no need to loop over each element. Also the `group_by()` operation
in dplyr offers a much more elegant way of applying a function repeatedly to
a group of data. Third, there are higher order functions like `map` (or
the tidyverse equivalent: `purrr::map`) that save you from loops most of the
time.

Anyway, here is the loop:

```{r }
for (i in 1:10) {
  print(i)
}
```

There will be one iteration of the loop for each component of the vector
`1:10`, with  `i` taking on the values of those components -- in the first
iteration, `i = (1:10)[1]`; in the second  iteration, `i = (1:10)[2]`; and so
on.

And we are not restricted to integer vectors, but can loop over any vector,
even over lists:

```{r }
z <- NULL
for (i in list(swiss, mtcars)) {
  z <- c(z, colnames(i))
}
z
```

But we saw a much clearer way of doing this above:

```{r }
unlist(map(list(swiss, mtcars), colnames))
```

